Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-02-06 paper-1/manuscript.tex
Rev intro

Existing models of DD focus on synaptic plasticity, either short-term \citep{Mill2011-ah, May2015-lt, Yarden2017-eh} or long-term \citep{Hertag2020-kc, Wacongne2012-ah}, and most models make use of strongly regimented network structure to harness lateral inhibition \citep{Garagnani2011-eu, Hertag2020-kc}, feedback inhibition \citep{Wacongne2012-ah}, or depression in feedforward pathways \citep{Mill2011-ah} as additional mechanisms to support DD. However, recent evidence from dissociated cell culture \citep{Kubota2021-dx} suggests that even randomly connected networks of cortical neurons are capable of deviance detection, implying that detailed network structure is not a requirement.

Neuronal adaptation, i.e., changes in intrinsic neuronal sensitivity as opposed to changes to synaptic function, is a common phenomenon in excitatory neurons \citep{Sanchez-Vives2000-df, OLeary2010-ar, Sanchez-Aguilera2014-fd}. While neuronal rate adaptation has been successfully used to model SSA \citep{Garagnani2011-eu}, it has been suggested that this mechanism is incapable of producing true deviance detection due to a lack of stimulus sensitivity \citep{Yarden2017-eh}. However, to the extent that individual neurons are sensitive to different stimuli, we argue that neuronal plasticity is equally well placed to produce DD as synaptic plasticity.

Here, we investigate how neuronal short-term plasticity, expressed as spike-triggered threshold adaptation, can lead to deviance detection. Specifically, we use a spatially constrained, randomly connected network of point neurons with threshold adaptation, and show that deviance detection emerges from this mechanism alone, requiring neither synaptic plasticity nor a tightly defined network structure. Furthermore, we model the conjunction of both neuronal and synaptic short-term plasticity, and study how these mechanisms interact within the network as it processes regular and irregular stimulus sequences.

>>>>

Existing models of DD place their primary focus on synaptic plasticity. While models using long-term plasticity have been proposed \citep{Wacongne2012-ah, Hertag2020-kc}, the rapid establishment of DD in vivo \citep{Taaseh2011-gg} suggests that fast-acting short-term plasticity mechanisms play a key role. Several models using synaptic short-term plasticity have been proposed \citep{Mill2011-ah, May2015-lt, Yarden2017-eh}, finding good agreement with physiological data.

Neuronal plasticity, on the other hand, has largely been neglected in previous work, used as a mere afterthought in the neuron model \citep{Mill2011-ah} or rejected a priori as incapable of producing DD \citep{Yarden2017-eh}. However, neuronal adaptation, i.e., an experience-dependent reduction in intrinsic neuronal sensitivity independent of synaptic function, is a common phenomenon in excitatory neurons \citep{Sanchez-Vives2000-df, Henze2001-xd, Sanchez-Aguilera2014-fd, Teeter2018-iz} that has been successfully used to model stimulus-specific adaptation (SSA), a precursor to and prerequisite for DD \citep{Garagnani2011-eu}. Neuronal adaptation is well placed to play a role in DD: It establishes quickly and recovers over the course of up to several seconds \citep{Henze2001-xd, Mensi2012-au}, allowing it to act as a memory of recent activity in a neuron.

Here, we investigate how neuronal threshold adaptation can lead to deviance detection. Specifically, we use a spatially constrained, randomly connected network of point neurons with spike-triggered threshold adaptation, and show that deviance detection emerges from this mechanism alone, requiring neither synaptic plasticity nor a tightly defined network structure. Furthermore, we model the conjunction of both neuronal and synaptic short-term plasticity, and study how these mechanisms interact within the network as it processes regular and irregular stimulus sequences.
